Program Transformations for miniKanren

### 2020-07-02 papers/aplas/evaluation.tex
Evaluation in progress

\section{Evaluation}
In our study we compared the CPD adaptation for \mk{} and the new non-conjunctive partial deduction.
We have also employed the branching heuristic instead of the deterministic unfolding in the CPD to check whether it can improve the quality of the specialization.

We used the following programs to test the specializers on.
\begin{itemize}
  \item Two implementations of an evaluator of logic formulas.
  \item Two implementations of a typechecker for a small expression language.
  \item A program to compute a unifier of two terms.
  \item A program to search for paths of a specific length in a graph.
\end{itemize}

The last two relations are described in~\cite{lozov2019relational} thus we will only briefly describe them here.

We focused on these particular examples because they are

\begin{itemize}
  \item Relational interpreters
  \item Natural
  \item Observable
\end{itemize}

% All these relations are relational interpreters.
% Their last argument is a boolean value which is $true$ if the other arguments are in relation and $false$ otherwise.

>>>>

\section{Evaluation}
In our study we compared the CPD adaptation for \mk{} and the new non-conjunctive partial deduction.
For some programs we have also employed the branching heuristic instead of the deterministic unfolding in the CPD to check whether it can improve the quality of the specialization.
We measured the execution time of each specialized program for some queries and compared it with the execution time of the original program.

We used the following programs to test the specializers on.
\begin{itemize}
  \item A program to compute a concatenation of three lists.
  \item A program to compute both the length of the list and its maximal element.
  \item Two implementations of an evaluator of logic formulas.
  \item Two implementations of a typechecker for a small expression language.
  \item A program to compute a unifier of two terms.
  \item A program to search for paths of a specific length in a graph.
\end{itemize}

The last two relations are described in~\cite{lozov2019relational} thus we will only briefly describe them in this paper.

We focused on these particular examples because they are examples of relational interpreters which are natural and observable.

\subsection{Concatenation of Three lists}

The relation \lstinline{doubleAppend$^o$} concatenates three lists by a conjunction of two calls of the \lstinline{append$^o$} relation (see fig.~\ref{doubleApp}).
These two calls share a variable --- an intermediate list \lstinline{ts} --- which should be treated with care during specialization.
Partial deduction does ignore the sharing of the variables which renders the approach to be unproductive for this kind of relations.
This is why this relation is particularly well known in the conjunctive partial deduction work.

The first list gets traversed twice to construct the result.
First, when the intermidiate list \lstinline{ts} is constructed during the first \lstinline{append$^o$} call and then, when \lstinline{ts} is read during the second call.
This double traversal negatively impacts the execution time.

\begin{figure*}[!h]
  \centering
  \begin{minipage}{0.8\textwidth}
    \begin{lstlisting}[label={doubleApp}, caption={Concatenation of three lists}, captionpos=b, frame=tb]
  let doubleAppend$^o$ xs ys zs res =
    fresh (ts) (
      append$^o$ xs ys ts /\ append$^o$ ts zs res)

  let rec append$^o$ xs ys zs = conde [
    (xs === [] /\ ys === zs);
    fresh (h t r) (
      xs === (h % t) /\
      zs === (h % r) /\
      append$^o$ t ys r)]
    \end{lstlisting}
  \end{minipage}
\end{figure*}

The better implementation of this relation (see fig.~\ref{doubleApp:etalon}) does not traverse the first list twice.
This etalon implementation is generated with the conjunctive partial deduction.

\begin{figure*}[!h]
  \centering
  \begin{minipage}{0.8\textwidth}
    \begin{lstlisting}[label={doubleApp:etalon}, caption={Etalon implementation of concatenation of three lists}, captionpos=b, frame=tb]
  let doubleAppend$^o$ xs ys zs res = conde [
    (xs === [] /\ append$^o$ ys zs res);
    fresh (h t r) (
      xs === h % t /\
      res === h % r /\
      doubleAppend$^o$ t ys zs r)]
    \end{lstlisting}
  \end{minipage}
\end{figure*}

We run all translated programs in the forward and the backward direction.
The forward direction \lstinline{doubleAppend$^o$ x y z ?} concatenates three given lists: the first one of length 700, the second and the third --- of length 1.
When run in the backward direction \lstinline{doubleAppend$^o$ ? ? ? r} , it searches for the first 100 list triples whose concatenation gives the given list of length 700.

There is no significant difference in the execution time in the backward direction between different specialized versions~\ref{tbl:doubleApp}.
However the execution time in the backward direction differs: the program generated by Ecce shows the same speedup as the etalon program, while non-conjunctive partial deduction slows the program down.
\todo{WHY?}

\begin{table}
  \centering
  \begin{tabular}{c||c||c}
                   & forward & backward \\
  \hline\hline
  Original         & 0.0040s & 0.0195s \\ \hline
  Ecce             & 0.0031s & 0.0201s \\ \hline
  Non-CPD          & 0.0055s & 0.0202s \\ \hline
  Etalon           & 0.0031s & 0.0203s \\
  \end{tabular}
  \caption{Evaluation results for doubleAppendo}
  \label{tbl:doubleApp}
\end{table}
